---
title: "Data Visualization Challenge"
subtitle: "R Setup in Cursor + Penguins and Simpson's Paradox"
format:
  html: default
  pdf: default
execute:
  echo: true
  eval: false
---

# 📈 Data Viz Challenge - Quick Start Guide

## Challenge Overview

In this challenge, you'll set up R inside Cursor (VS Code–compatible), get comfortable with terminals, shells, and command lines, and then prepare to build a two-panel or two-facet ggplot using the Palmer Penguins dataset to illustrate Simpson's paradox.

By the end of Part 1, you will be able to run R code from Cursor and render R chunks in Quarto. Plotting tasks are scaffolded with placeholders you will complete later.

## Part 1: R in Cursor Setup ⚙️

Follow these steps (adapted from the official R extension guidance) to get R working smoothly in Cursor.

### 1. Install R

- Download and install R (version ≥ 3.4.0) from [CRAN R downloads](https://cran.r-project.org/)
- During installation on Windows, select the option to **Write R path to the registry** (recommended)

### 2. Install the R extension in Cursor

- Open Extensions: `Ctrl+Shift+X`
- Search for "R" (publisher: REditorSupport) and install the extension named "R"
- Also install this/these extensions if you haven't already:
  - "Quarto" (for .qmd editing and preview)

### 3. Install languageserver (enables R language features)

You can install directly from an R session. In Cursor:

- Open Command Palette: `Ctrl+Shift+P` → type "R: Create R Terminal" → Enter
- In the R console that appears, run:

```r
install.packages("languageserver")
```

::: {.callout-note}
Cursor may prompt to install `languageserver` for you automatically; accepting that prompt is fine.
:::

### 4. Recommended tools for better plot viewing

For improved ggplot visualization that scales with your screen:

**Option 1: httpgd (Recommended)**
```r
install.packages("httpgd")
```
- Creates a web-based plot viewer that opens in your browser
- Scalable, interactive, and high-quality SVG plots
- Works great with the R extension

**Option 2: R Plot Viewer Extension**
- In Cursor: `Ctrl+Shift+X` → search for "R Plot Viewer" 
- Provides a dedicated plot panel with better scaling

**Option 3: External viewers**
```r
# For Windows - opens plots in external viewer
windows()  # before creating plots

# For macOS
quartz()

# For Linux
x11()
```

::: {.callout-tip}
After installing `httpgd`, restart your R session and create a plot. The R extension should automatically use the web-based viewer for better scaling and quality.
:::

### 5. Verify R works in Cursor

1) Create a new file `hello.R` and add a couple lines:

```r
summary(cars)
mean(cars$speed)
```

2) With the editor caret on a line, press `Ctrl+Enter` to send the current line to the active R terminal.

3) If Cursor can't find R automatically on Windows, set the R path:
   - Open Settings: `Ctrl+,`
   - Search for "Rterm"
   - Set `R › Rterm: Windows` to something like `C:\\Program Files\\R\\R-4.4.1\\bin\\x64\\R.exe` (adjust version/path as installed)

### 6. Running R chunks in Quarto (.qmd)

With the R extension (and Quarto extension installed), you can execute R code chunks inside Quarto documents.

- Create or open a `.qmd` file and insert an R chunk:

```{r}
#| label: setup
#| message: false
#| warning: false
R.version.string
```

- Use the chunk toolbar (Run/Run Above/Run All) to execute. If the toolbar is hidden, place the cursor in the chunk and use `Ctrl+Shift+P` → "Quarto: Run Cell" or send lines with `Ctrl+Enter`.

---

## Part 2: Terminals, Shells, and Command Lines 🖥️

🧩 **Putting It All Together**

Here's the hierarchy that helps when switching between the system terminal and the R terminal:

- **Terminal** → the app that gives you a text window (e.g., Cursor's Integrated Terminal)
- **Shell** → the program that runs inside that window
  - Examples (Windows): PowerShell, `cmd.exe`
  - Examples (macOS/Linux): `bash`, `zsh`
- **Command line** → the interaction style: typing commands into the shell
- **R Terminal** → a specialized program you run from the shell, which then gives you an R-only command line

**In Cursor:**
- Open the integrated terminal: `Ctrl+`` (backtick) - this gives you the system shell
- Create an R terminal: `Ctrl+Shift+P` → "R: Create R Terminal" - this runs R inside the terminal
- Use the R terminal for R commands; use the system shell (PowerShell/bash) for tools like `git` and `quarto`

---

## Part 3: Penguins + Simpson's Paradox (Placeholders) 🐧

You'll use Palmer Penguins to demonstrate how a relationship can reverse when stratified (Simpson's paradox). For now, set up the project and confirm packages install.

### A. Install and load packages

```{r}
install.packages(c("palmerpenguins", "ggplot2"))

library(palmerpenguins)
library(ggplot2)
data(penguins)
```

### B. Single overall trend (placeholder)

```{r}
ggplot(data = penguins,
        aes(x = bill_length_mm,
        y = bill_depth_mm)) +
  theme_minimal(32) +
  geom_point() +
  scale_color_manual(values = c("darkorange","purple","cyan4")) +
  labs(title = "Penguin bill dimensions (omit species)",
       subtitle = "Palmer Station LTER",
       x = "Bill length (mm)",
       y = "Bill depth (mm)") +
  theme(plot.title.position = "plot",
        plot.caption = element_text(hjust = 0, face= "italic"),
        plot.caption.position = "plot") +
  geom_smooth(method = "lm", se = FALSE, color = "gray50")
```

### C. Stratified view via facets (canonical)

```r
# TODO: Facet by species (or island) to reveal group-wise trends
# ggplot(penguins, aes(x = bill_length_mm, y = flipper_length_mm, color = species)) +
#   geom_point(alpha = 0.7) +
#   geom_smooth(method = "lm", se = FALSE) +
#   facet_wrap(~ species)
```

### D. Two-panel composition (canonical alternatives)

```r
# TODO: Show overall vs stratified side-by-side
# Option 1: Use facets as above (keeps in one ggplot)
# Option 2: Create two separate ggplots and combine with patchwork or gridExtra

# install.packages("patchwork")
# library(patchwork)

# p_overall <- ggplot(...) + ...
# p_facet   <- ggplot(...) + ... + facet_wrap(~ species)
# p_overall + p_facet
```

### E. Brief write-up (placeholder)

- Explain how the overall relationship differs from the within-group relationships
- Identify the grouping variable that flips or alters the trend (species or island)

---

## Submission Checklist ✅

- [ ] R installed (Windows: path written to registry)
- [ ] R extension installed in Cursor (plus Quarto extension)
- [ ] `languageserver` installed
- [ ] `httpgd` installed (recommended)
- [ ] Able to send lines from `hello.R` to the R terminal
- [ ] Able to run an R chunk inside a `.qmd`
- [ ] Packages `palmerpenguins` and `ggplot2` installed
- [ ] Placeholder plots and notes added to this document

---

## Need Help? 🆘

- R extension documentation: [VS Code R Extension Wiki](https://github.com/REditorSupport/vscode-R/wiki)
- Quarto docs: [Quarto Guide](https://quarto.org/docs)
- General tip: search for "VS Code" results—most answers apply to Cursor as well


